A data-driven model of nucleosynthesis with chemical tagging in a lower-dimensional latent space

### 2019-09-17 article/mcfa.tex
add schematic

@@

@@

\begin{eqnarray}
	I(\scoremeans,\scorecovs) &=& -\sum_{\numcomponents=1}^{\NumComponents}\log{p(\scoremeans_k,\scorecovs_k)} + \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log{|\mathcal{F}(\scoremeans_k,\scorecovs_k)|} \nonumber \\
	&=& \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log\left[(\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+3)}2^{-\NumData}|\scorecovs_k|^{-(\NumData + 2)}\right] \nonumber \\
	&& \cdots \,\, -\sum_{\numcomponents=1}^{\NumComponents}\log{|\scorecovs_k|}^{\frac{1}{2}(\NumLatentFactors + 1)} \\
I(\scoremeans,\scorecovs) &=& \frac{1}{4}\NumLatentFactors(\NumLatentFactors+3)\sum_{\numcomponents=1}^\NumComponents\log{\NumData\weight_k} - \frac{KD}{2}\log{2} \nonumber \\ 
&& \cdots \,\, -\frac{1}{2}(2\NumLatentFactors+3)\sum_{k=1}^{K}\log{|\scorecovs_k|}  \quad . \label{eq:prior_xi_omega} 
\end{eqnarray}

@@

\begin{widetext}
\noindent{}Thus the message length to encode $\factorloads$ is given by
\begin{eqnarray}
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \quad . \label{eq:prior_L}
\end{eqnarray}

\noindent{}Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
\begin{eqnarray}
	I(\vec\Psi, \vec\data) &=& -\log\likelihood(\vec\data|\vec\Psi)
 +\frac{1}{4}\left(\NumLatentFactors + 4\right)\left(\NumLatentFactors - 1\right)\sum_{\numcomponents=1}^\NumComponents\log\weight_\numcomponents + \left(\NumComponents - \frac{1}{2}\right)\log{\NumData}
 +\frac{1}{2}\NumDimensions\log|\textrm{Cov}\left(\factorloads\transpose\right)| \nonumber \\
  && \cdots -\frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| + \textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents|  -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2}  \quad . \label{eq:message-length}
 \end{eqnarray}
\end{widetext}

@@

	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-1-gridsearch-mml-contours.pdf} \\
	\end{tabular}
	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-2-gridsearch-mml-contours.pdf} \\
	\end{tabular}
    \caption{Metrics from our grid search for the toy model for two sample sizes: 
    		 $N = 100,000$ (left) and $N = 5,000$ (right). 
    		 The top panels show the negative log likelihood
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors 
			 $\NumLatentFactors$ and number of clusters 
			 $\NumComponents$ using the generated data in our toy model.
			 The middle panels shows 
			 the BIC (Eq.~\ref{eq:bic}) for those 
			 combinations, and the lower panel shows the 
			 message length. The white marker indicates the
			 lowest value in each panel. A line connects to the true value (black point) to guide the eye.}
    \label{fig:experiment-1-gridsearch}
\end{figure*}

@@

\begin{figure*}[t!]
	%\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-1-compare-all.pdf}
	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
    		 and specific variances $\specificvariance$ (right) compared to the 
		 	 true data generating values
		 	 for Experiment~1 (Section~\ref{sec:exp-1}). The agreement is excellent.}
    \label{fig:exp1-compare}
\end{figure*}

@@

>>>>

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/schematic.png}
	\caption{A schematic that visualises the model components.
			 The data are shown on the left in greyscale, and magnified on the
			 right where each star (row) is coloured by its identified
			 component in latent space.
			 For each chemical abundance (column) there is a mean value and
			 variance that is independent of the latent space.
			 The latent factors $\factorloads$ are analogous to nucleosynthetic
			 yields and are common to all stars. The factor scores $\factorscores$
			 have an entry for each yield, for each star (row). The latent scores are modelled
			 by a mix of multivariate normal distributions of $\NumComponents$
			 components, which are coloured accordingly. The matrices of factor 
			 scores $\factorscores$ and factor loads $\factorloads$ are 
			 visualised at the bottom. For clarity here we show transposed
			 matrices (e.g., see Eq. \ref{eq:generative-model}).}
	\label{fig:schematic}
\end{figure*}

@@

\vspace{1em}

@@

\begin{eqnarray}
	I(\scoremeans,\scorecovs) &=& -\sum_{\numcomponents=1}^{\NumComponents}\log{p(\scoremeans_k,\scorecovs_k)} + \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log{|\mathcal{F}(\scoremeans_k,\scorecovs_k)|} \nonumber \\
%	&=& \frac{1}{2}\sum_{\numcomponents=1}^{\NumComponents}\log\left[(\NumData\weight_k)^{\frac{1}{2}\NumLatentFactors(\NumLatentFactors+3)}2^{-\NumData}|\scorecovs_k|^{-(\NumData + 2)}\right] \nonumber \\
%	&& \cdots \,\, -\sum_{\numcomponents=1}^{\NumComponents}\log{|\scorecovs_k|}^{\frac{1}{2}(\NumLatentFactors + 1)} \\
I(\scoremeans,\scorecovs) &=& \frac{1}{4}\NumLatentFactors(\NumLatentFactors+3)\sum_{\numcomponents=1}^\NumComponents\log{\NumData\weight_k} - \frac{KD}{2}\log{2} \nonumber \\ 
&& \cdots \,\, -\frac{1}{2}(2\NumLatentFactors+3)\sum_{k=1}^{K}\log{|\scorecovs_k|}  \quad . \label{eq:prior_xi_omega} 
\end{eqnarray}

@@

\noindent{}Thus the message length to encode $\factorloads$ is given by
\begin{widetext}
\begin{eqnarray}
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \quad . \label{eq:prior_L}
\end{eqnarray}
%\end{widetext}
Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
%\begin{widetext}
\begin{eqnarray}
	I(\vec\Psi, \vec\data) &=& -\log\likelihood(\vec\data|\vec\Psi)
 +\frac{1}{4}\left(\NumLatentFactors + 4\right)\left(\NumLatentFactors - 1\right)\sum_{\numcomponents=1}^\NumComponents\log\weight_\numcomponents + \left(\NumComponents - \frac{1}{2}\right)\log{\NumData}
 +\frac{1}{2}\NumDimensions\log|\textrm{Cov}\left(\factorloads\transpose\right)| \nonumber \\
  && \cdots -\frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| + \textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents|  -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2}  \quad . \label{eq:message-length}
 \end{eqnarray}
\end{widetext}

%\hphantom{asdfasdf\\ asdf \\}
%\vspace{5em}
%\hphantom{asdfasdf\\ asdf \\}
%\vspace{5em}

%\newpage

%\vspace{2em}

@@

	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-1-gridsearch-mml-contours.pdf} \\
	\end{tabular}
	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-2-gridsearch-mml-contours.pdf} \\
	\end{tabular}
    \caption{Metrics from our grid search for the toy model for two sample sizes: 
    		 $N = 100,000$ (left) and $N = 5,000$ (right). 
    		 The top panels show the negative log likelihood
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors 
			 $\NumLatentFactors$ and number of clusters 
			 $\NumComponents$ using the generated data in our toy model.
			 The middle panels shows 
			 the BIC (Eq.~\ref{eq:bic}) for those 
			 combinations, and the lower panel shows the 
			 message length. The white marker indicates the
			 lowest value in each panel. A line connects to the true value (black point) to guide the eye.}
    \label{fig:experiment-1-gridsearch}
\end{figure*}
\vspace{-50em}

@@

@@

\begin{figure*}
	%\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-1-compare-all.pdf}
	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
    		 and specific variances $\specificvariance$ (right) compared to the 
		 	 true data generating values
		 	 for Experiment~1 (Section~\ref{sec:exp-1}). The agreement is excellent.}
    \label{fig:exp1-compare}
\end{figure*}
